How to Deal With Measurement Error in Measurement Models

### 2023-05-19 paper.qmd
beginning to remove BISG stuff

```{=tex}
\usetikzlibrary{positioning}
\tikzset{
    > = stealth,
    every node/.append style = {
        draw = none
    },
    every path/.append style = {
        arrows = ->
    },
    hidden/.style = {
        draw = black,
        shape = circle,
        inner sep = 1pt
    }
}
```

# Introduction

@@

In practice, information about the latent variable is often lost when researchers move from measurement to theory-testing. Measurement models do not simply output a single value for the underlying latent variable. Instead, by virtue of being statistical models, they produce *estimates of uncertainty* for each observation. This is particularly true for Bayesian measurement models, whose output is the full posterior distribution of latent variable values according their relative plausibility---not a single point estimate. Failure to propagate this uncertainty from the measurement model into the theory-testing model, as I will show, can lead to biased and/or overly confident substantive conclusions.

In this paper I develop two methods---one for continuous latent variables and one for discrete latent variables---which incorporate measurement model uncertainty into theory-testing models. I will use two running examples to help illustrate these methods. The first is the Bayesian Item-Response Theory (IRT) model used to measure political ideology [@clinton2004]. Political ideology is typically imagined as existing on a latent left-right dimension. Bayesian IRT measurement models produce a posterior distribution of continuous values for each actor (e.g. member of Congress) on the left-right scale. The second illustrative model is Bayesian Improved Surname Geocoding (BISG), used to measure individuals' race and ethnicity in the absence of self-reported values for these characteristics [@elliott2009]. Unlike political ideology, which could conceivably take any real-number value on the left-right scale, race and ethnicity are usually considered to be discrete, un-ordered, categories.[^1] BISG measurement models, therefore, output a simplex for each individual corresponding to the posterior probability that they identify as a member of each potential racial or ethnic group.

[^1]: For an alternative conception of race in social identity theory see @abdelal2006 .

@@

@@

I will demonstrate my argument using the causal graph framework [@pearl2000]. Causal graphs are heuristic tools that map out causal relationships between variables in a particular system. Each node represents a variable, and the directed edges between nodes represent hypothesized causal impacts of one variable on another. These directed-acyclic-graphs (DAGs) are useful because they allow us to determine the set of variables we need to condition on in order to get an unbiased estimate of the effect of our primary independent variable on the dependent variable. This set of confounders is defined by the variables which are needed to close every "backdoor" path between the primary independent and dependent variables.[^2]

[^2]: See @cinelli2020 for a more complete introduction to deconfounding using DAGs.

@@

sim_cont_data(500, noise_x = 5, noise_a = 0)$dat |> 
  arrange(x_meas) |> 
  mutate(id = row_number()) |> 
  ggplot(aes(x = x_meas, y = id)) +
  geom_pointrange(aes(xmin = x_meas - x_sd, xmax = x_meas + x_sd), 
                  alpha = .35, size = .25, color = met.brewer("Isfahan1", 1)) +
  geom_pointrange(aes(xmin = x_meas - x_sd * 1.96, xmax = x_meas + x_sd * 1.96), 
                  alpha = .1, size = .25, color = met.brewer("Isfahan1", 1)) +
  # geom_point(color = "red", size = .3, alpha = .2) + 
  labs(x = expression(paste("Estimated Ideal Point, ", theta^{`*`})), y = "Groups (sorted)",
       caption = "Means and Standard Errors") +
  theme_ggdist() +
  theme(text = element_text(family = "serif"))
```

@@

While there may exist some backdoor paths through $e^{\theta}$ and $P$ in a hypothetical theory-testing model, I will assume that the outcome of interest, $Y$ is unaffected by anything other than the direct causal effect $\theta^* \longrightarrow Y$.[^3] The purpose of this simplification is to highlight the consequences of random measurement error during parameter estimation of a theory-testing model. Using the generative causal model in @fig-ideal-dag-theory we can simulate data with a known parameter for the effect of group ideal point on the outcome $Y$. Then we fit two linear regression models to estimate this parameter, $\beta$. @eq-cont-no-me-model is the naive theory-testing model where $X_{\text{MEAS},i}$ corresponds to a group's mean ideal point estimate from the Bayesian IRT measurement model (see right panel of @fig-measurement-model). This is in contrast to the joint measurement theory-testing model in @eq-cont-me-model, which models $X_{\text{MEAS},i}$ as an outcome of $X_{\text{TRUE},i}$ (an unobserved parameter for each observation)[^4] and $X_{\text{SE},i}$ which is the observed standard deviation of the posterior distribution for each groups' ideal point. The estimates of $X_{\text{TRUE},i}$ are then used in the linear model which predicts the outcome $y_i$.

[^3]: In principle, we could close any backdoor paths through $P$ by conditioning on it directly because the level of group participation is directly observed.

[^4]: The parameters $X_{\text{TRUE},i}$ should not be confused with the true ideal points produced in the simulation. These true values are never seen by either model.

>>>>

```{=tex}
\usetikzlibrary{positioning}
\tikzset{
    > = stealth,
    every node/.append style = {
        draw = none
    },
    every path/.append style = {
        arrows = ->
    },
    hidden/.style = {
        draw = black,
        shape = circle,
        inner sep = 1pt
    }
}
```
# Introduction

@@

In practice, information about the latent variable is often lost when researchers move from measurement to theory-testing. Measurement models do not simply output a single value for the underlying latent variable. Instead, by virtue of being statistical models, they produce *estimates of uncertainty* for each observation. This is particularly true for Bayesian measurement models, whose output is the full posterior distribution of latent variable values according their relative plausibility---not a single point estimate. Failure to propagate this uncertainty from the measurement model into the theory-testing model, as I will show, can lead to mistaken conclusions regarding the underlying research question. And unlike so-called "classical" measurement error, whose attenuation bias properties are generally well understood, the mistakes I investigate can lead to bias in unpredictable directions.

In this paper I demonstrate the problems associated with failing to include measurement model measurement error in theory-testing models, and I develop a method for overcoming these issues. This method is appropriate for any continuous-valued latent variable generated from a Bayesian measurement model. One of the most common types of these in the political science literature is the Bayesian Item-Response Theory (IRT) model used to measure political ideology [@clinton2004; @bafumi2005]. Political ideology is typically imagined as existing on a latent left-right dimension. Bayesian IRT measurement models produce a posterior distribution of continuous values for each actor (e.g. member of Congress) on the left-right scale.

@@

```{r}
#| label: fig-measurement-model
#| fig-cap: "Ignoring Measurement Error in Measurement Models"
N <- 100000
tibble(x = c(rnorm(N, sd = 1.5),
             rnorm(N, sd = 4),
             sn::rsn(N, alpha = -4, omega = 4, xi = 3)),
       y = c(rep("a", N), rep("b", N), rep("c", N))) |> 
  ggplot(aes(x = x)) +
  geom_density(color = "black", fill = "darkcyan", alpha = .5) +
  facet_wrap(~ y, nrow = 1) +
  theme_void()
```

@@

I will demonstrate my argument using the causal graph framework [@pearl2000]. Causal graphs are heuristic tools that map out causal relationships between variables in a particular system. Each node represents a variable, and the directed edges between nodes represent hypothesized causal impacts of one variable on another. These directed-acyclic-graphs (DAGs) are useful because they allow us to determine the set of variables we need to condition on in order to get an unbiased estimate of the effect of our primary independent variable on the dependent variable. This set of confounders is defined by the variables which are needed to close every "backdoor" path between the primary independent and dependent variables.[^1]

[^1]: See @cinelli2020 for a more complete introduction to deconfounding using DAGs.

@@

sim_cont_data(500, noise_a = 2, true_b = 1)$dat |> 
  arrange(x_meas) |> 
  mutate(id = row_number()) |> 
  ggplot(aes(x = x_meas, y = id)) +
  geom_pointrange(aes(xmin = x_meas - x_sd, xmax = x_meas + x_sd), 
                  alpha = .35, size = .25, color = met.brewer("Isfahan1", 1)) +
  geom_pointrange(aes(xmin = x_meas - x_sd * 1.96, xmax = x_meas + x_sd * 1.96), 
                  alpha = .1, size = .25, color = met.brewer("Isfahan1", 1)) +
  # geom_point(color = "red", size = .3, alpha = .2) + 
  labs(x = expression(paste("Estimated Ideal Point, ", theta^{`*`})), y = "Groups (sorted)",
       caption = "Means and Standard Errors") +
  theme_ggdist() +
  theme(text = element_text(family = "serif"))
```

@@

While there may exist some backdoor paths through $e^{\theta}$ and $P$ in a hypothetical theory-testing model, I will assume that the outcome of interest, $Y$ is unaffected by anything other than the direct causal effect $\theta^* \longrightarrow Y$.[^2] The purpose of this simplification is to highlight the consequences of random measurement error during parameter estimation of a theory-testing model. Using the generative causal model in @fig-ideal-dag-theory we can simulate data with a known parameter for the effect of group ideal point on the outcome $Y$. Then we fit two linear regression models to estimate this parameter, $\beta$. @eq-cont-no-me-model is the naive theory-testing model where $X_{\text{MEAS},i}$ corresponds to a group's mean ideal point estimate from the Bayesian IRT measurement model (see right panel of @fig-measurement-model). This is in contrast to the joint measurement theory-testing model in @eq-cont-me-model, which models $X_{\text{MEAS},i}$ as an outcome of $X_{\text{TRUE},i}$ (an unobserved parameter for each observation)[^3] and $X_{\text{SE},i}$ which is the observed standard deviation of the posterior distribution for each groups' ideal point. The estimates of $X_{\text{TRUE},i}$ are then used in the linear model which predicts the outcome $y_i$.

[^2]: In principle, we could close any backdoor paths through $P$ by conditioning on it directly because the level of group participation is directly observed.

[^3]: The parameters $X_{\text{TRUE},i}$ should not be confused with the true ideal points produced in the simulation. These true values are never seen by either model.
